---
title: "Jail feminism"
author: "Erzhan Umarov"
date: '2022-12-11'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)

states<-read.csv("data/state_names_and_codes.csv")
# df<-read.csv("data/incarceration_trends.csv")
# saveRDS(df,"data/df.Rds")

# Savind file as Rds to load faster
df<-readRDS("data/df.Rds")

source("analysis.R")
source("a4-helpers.R")
```


<!-- Section 1 -->
## Introduction

The Objective of this report is to identify and expose the patterns of inequality in incarnation data. 

In this data set we'll be analyzing the black population and white population in jails and the relevant trends from 1988 to 2018.


<!-- Section 2 -->
## Data Summary

```{r echo=FALSE}
# Defining the variable
var<-"white_jail_pop"

# Generating max min county names for variable
max2018<-get_max_county(df=df,variable=var,yearx=2018)
min2018<-get_min_county(df=df,variable=var,yearx=2018)

# geting avarage for the variable in 1988 and 2018
avg1988<-get_county_wise_avg(df=df,variable=var,yearx=1988)
avg2018<-get_county_wise_avg(df=df,variable=var,yearx=2018)

# geting avarage for the total population in 1988 and 2018
avgToal1988<-get_county_wise_avg(df=df,variable="total_jail_pop",yearx=1988)
avgToal2018<-get_county_wise_avg(df=df,variable='total_jail_pop',yearx=2018)

k<-avgToal2018-avgToal1988<avg2018-avg1988

change<-if(!k){"The white population has been grown more than the overall avarage in prisons and thus there's an trend of increasing white population."}else{"The white population has been grown less than the overall avarage in prisons and thus there's an trend of decreasing white population.."}
```


In 2018, The county with most of the white population in jails is `r max2018`and the county with lowest white population in jail is `r min2018`

in 1988, the white population in jails were `r avg1988` and in 2018 it was `r avg2018` which is a growth of `r round((avg2018-avg1988)/avg1988*100,2)`%


As per the total population in jails, in 1988 `r avgToal1988` and `r avgToal2018` in 2018, which is a growth of `r round((avgToal2018-avgToal1988)/avgToal1988*100,2)`%

Considering the fact, it can be identified that `r change`

<!-- Section 3 -->
## Growth of U.S Prison Population

```{r echo=FALSE}
plot_jail_pop_for_us()

```

As it indicates there has been a huge increase of population in jail from 1980 to 2008 and then it indicates a pattern of reducing or fixating as a total in all county.

<!-- Section 4 -->
## Growth of Prison Population by state

```{r echo=FALSE}
plot_jail_pop_by_state(states = unique(df$state)[1:5])


```

As it indicates, CA has a tremendoues increase of the population 1980 to 1990 and seems to become saturated. all the other states does not show a big variation but is shows a groth of the population and AK seems to have no population.


<!-- Section 5 -->
## Comparisson of black and white population in Jails

```{r echo=FALSE}
get_wb_plot(df)
```


```{r echo=FALSE}

k<-get_bw_ratio(df)%>%
  filter(year==2018)%>%
  head(10)
statesx<-unique(k$state)[1:5]
```

As it indicates, by 2018, top 10 places where the black/white jail population has extremes are as below.

```{r echo=FALSE}
knitr::kable(k)
```

For these top 5 states, the ration has been changed as below figure with time.

```{r echo=FALSE}

k<-get_bw_ratio(df)%>%
  filter(state%in%statesx)

k<-k%>%
  tidyr::spread(key = state,value = ratio)

k<-melt(k,id="year")

ggplot(k,            
         aes(x = year,
             y = value,
             color = variable)) +  geom_line() +
  labs(title = "Black/White Jail population yearly", caption = "Black/ White jail population is yearly plotted for top 5 states with ratio extreme.")

```

as it indicates, by far DC has an extreme of black jail population and it has been in its highest and has came back but yet it is way higher than other states.


<!-- Section 6 -->
## Potential inequality map of Black/White Ratio

```{r echo=FALSE}
toMap<-get_bw_ratio(df)%>%
  filter(year==2018)

get_bw_map(df = toMap)


```